---
title: "Homework 1 - The Learning Procedure"
author: "Parsa Delavari"
date: "Due 23 September at 11pm"
output:
  bookdown::pdf_document2:
    includes:
      in_header: preamble.tex
    number_sections: true
    toc: FALSE
urlcolour: blue
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, error = TRUE)
library(tidyverse)
library(testthat)
library(Stat406)
data("bccovid_train")
data("bccovid_test")
data("bcpop")
theme_set(theme_bw()) # Remove the ggplot grey background, isn't it ugly?
```

### Instructions {.unnumbered}

-   Use the space inside of

```         
  ::: {.solbox data-latex=""}


  :::
```

to answer the following questions.

-   Do not move this file or copy it and work elsewhere. Work in the
    same directory.
-   Use a new branch named whatever you want. Create it now! Can't come
    up with something, try
    [here](https://alator21.github.io/repository-name-generator/). Make
    a small change, say by adding your name. Commit and push now!
-   Try to Knit your file now. Are there errors? Fix them now, not at
    11pm on the due date.
-   There MUST be some text between `::: {.solbox}` and the next `:::`
    or this will fail to Knit.
-   If your code or figures run off the edge of the `.pdf`, you'll lose
    2 points automatically.
-   Be sure to add your name in the `author` field at the top.

::: {.solbox data-latex=""}
**Generative AI Self Report**

If you use any generative AI tool on homeworks or labs, indicate here
how you used the tools and how they contributed to your learning/ability
to understand the assignment. Include every prompt that you used.

This self report will not affect your grade in any way. Please be as
honest as possible.

-   your response here \*
:::

\clearpage

# Forecasting COVID in BC

In the course `Stat406` package, you will find COVID-19 case data for
the five health authorities in BC. We're going to play with a few models
for making forecasts. Your goal is to **predict 2-week ahead case counts
for all 5 health authorities**.

There are 2 BC Covid datasets we need for this assignment:
`bccovid_train` and `bccovid_test`. Both have 6 variables: `HA`, `date`,
`cases+14`, `cases+0`, `cases-7`, `cases-14`. `cases+14` is the
response: it is the total number of cases observed in that specific HA,
in the week ending 14 days after `date`. The other `cases` variables are
features: the total number of cases in the week ending on the date,
those for 1 week earlier and those for 2 weeks earlier. For complete
descriptions, try `?bccovid` after loading the course package (or
examine the package documentation on the
[web](https://ubc-stat.github.io/stat-406-rpackage/reference/bccovid.html)).

## (1 point)

Using only the training data: a. (0.5 point) make a plot of today's
cases over time. colour the points by health authority. b. (0.5 point)
Write a few sentences discussing the data. Describe any similarities or
differences across health authorities you notice. How do cases seem to
"evolve" over time?

> **Hint:** to use these covariates in `ggplot()`, you have to put them
> inside "backtick" marks, like `` `cases+0` ``.

::: {.solbox data-latex=""}
a.  

```{r q1-solution}
ggplot(bccovid_train) +
  geom_point(alpha = 0.6, aes(x = date, y = `cases+0`, colour = HA)) +
  labs(title = "Weekly COVID-19 Cases in BC Health Authorities",
       x = "Date",
       y = "Cases (week ending on date)",
       colour = "Health Authority") 
```

b.  
:::

\clearpage

## (2 points)

Estimate a linear model of 2-week ahead cases (`cases+14`) on the three
other case features using the training data.

a.  (0.25 points) Plot the residuals against time (again coloured by
    HA).
b.  (0.5 points) Make a QQ-plot of the residuals (coloured by HA). Do
    these plots seem support the standard assumptions for a linear
    model? Why or why not?
c.  (0.5 point) Report the slope coefficient on `cases+0`. Describe what
    this coefficient means in the context of this problem. Does this
    interpretation make sense, why or why not?
d.  (0.75 points) Does this model seem appropriate? Justify your answer.

**Hint:** to use these covariates in a formula, you have to put them
inside "backtick" marks, like `` `cases+0` ``.

::: {.solbox data-latex=""}
a.  

```{r q2-solution}
# Fit linear model
lm_fit <- lm(`cases+14` ~ `cases+0` + `cases-7` + `cases-14`, data = bccovid_train)

length(residuals(lm_fit))

bccovid_train$resid <- residuals(lm_fit)

ggplot(bccovid_train, aes(x = date, y = resid, colour = HA)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Time", y = "Residuals") +
  theme_minimal()
```

b.  

```{r q2-solb}
ggplot(bccovid_train, aes(sample = resid, colour = HA)) +
  stat_qq(alpha = 0.6) +
  labs(title = "QQ Plot of Residuals") +
  theme_minimal()

coef(lm_fit)["`cases+0`"]

```

c.  

d.  

e.  
:::

\clearpage

## (2 points)

Examine the `bcpop` data set. Using an appropriate `*_join()`, add this
column to the training data. Now scale all the columns by the population
to get "cases per capita". Multiply by 100,000 just to get reasonable
looking numbers.

-   (0.5 point) Make a plot of today's cases over time scaled by
    population (and coloured by HA). Furthermore, scale the y-axis on
    the `log10()` scale. (Hint: try `?scale_y_log10`). How does this
    compare to the plot in Question 1.1? Write 2 complete sentences.
-   (1.5 points) Repeat Question 1.2 with both your scaled data after
    taking `log()` of all the predictors AND using HA as a categorical
    predictor.

::: {.solbox data-latex=""}
**First bullet.**

```{r q3-solution}

bccovid_train2 <- bccovid_train %>%
  left_join(bcpop, by = "HA")

bccovid_train2

bccovid_train2 <- bccovid_train2 %>%
  mutate(`cases+0_pc`  = (`cases+0`  / pop) * 100000,
         `cases-7_pc`  = (`cases-7`  / pop) * 100000,
         `cases-14_pc` = (`cases-14` / pop) * 100000,
         `cases+14_pc` = (`cases+14` / pop) * 100000)

ggplot(bccovid_train2, aes(x = date, y = `cases+0_pc`, colour = HA)) +
  geom_point(alpha = 0.6) +
  scale_y_log10() +
  labs(title = "Per-capita weekly cases (per 100,000) in BC",
       y = "Cases per 100,000 (log scale)",
       x = "Date")
```

\clearpage

**Repeat Q2 with different data.**

a.  

```{r q3-solb}
lm_fit2 <- lm(log(`cases+14_pc`) ~ log(`cases+0_pc`) +
                                     log(`cases-7_pc`) +
                                     log(`cases-14_pc`) +
                                     HA,
              data = bccovid_train2)

bccovid_train2$resid2 <- residuals(lm_fit2)

# Residuals vs time
ggplot(bccovid_train2, aes(x = date, y = resid2, colour = HA)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Time (scaled + log predictors)")

# QQ plot
ggplot(bccovid_train2, aes(sample = resid2, colour = HA)) +
  stat_qq(alpha = 0.6) +
  stat_qq_line() +
  labs(title = "QQ Plot of Residuals (scaled + log predictors)")
```

b.  

c.  

d.  

e.  
:::

## (1 point)

Plot the fitted values from the two models against each other (fits from
the Question 1.2 on the x-axis, with fits from Question 1.3 on the
y-axis). Be sure to transform them as needed to be comparable. Colour
the points by HA. What do you notice?

::: {.solbox data-latex=""}
```{r q4-solution}
# fitted values
bccovid_train2$fit1 <- fitted(lm_fit)   # from Q1.2 (raw counts)

bccovid_train2$fit2 <- exp(fitted(lm_fit2))   # from Q1.3 (log per capita)
bccovid_train2$fit1_pc <- (bccovid_train2$fit1 / bccovid_train2$pop) * 100000

ggplot(bccovid_train2, aes(x = fit1_pc, y = fit2, colour = HA)) +
  geom_point(alpha = 0.6) +
  labs(title = "Fitted values: raw vs log-scaled models",
       x = "Fitted per-capita cases (Model 1)",
       y = "Fitted per-capita cases (Model 2)")

```

(placeholder text)
:::

\clearpage

## (1 point)

Calculate the root mean squared error of both models. Discuss the
results in 2 sentences.

::: {.solbox data-latex=""}
```{r q5-sol-table}
rmse1 <- sqrt(mean(residuals(lm_fit)^2))

# observed per-capita
obs_pc <- (bccovid_train2$`cases+14` / bccovid_train2$pop) * 100000

# fitted per-capita (already back-transformed earlier)
fit2 <- exp(fitted(lm_fit2))

rmse2 <- sqrt(mean((obs_pc - fit2)^2))


print(rmse1)
print(rmse2)

```

(placeholder text)
:::

## (1 point)

Use the K-fold CV function from class (loaded below) to compare these
two models (the first model being the one with `bccovid_train` from
Question 1.2, and the second being the one with the log of case rates
and the factor for HA from Question 1.3). Use $K = 10$. In order to do
this, you'll need to be sure that the errors are on the same scale.

Use `Cases` as the scale. This means you'll have to specify the model
and the error function carefully.

a.  (0.5 points) What are the CV scores for both models?
b.  (0.25 points) Which do you prefer?
c.  (0.25 points) How much more accurate is it?

```{r kfoldcv}
#' @param data The full data set
#' @param estimator Function. Has 1 argument (some data) and fits a model.
#' @param predictor Function. Has 2 args (the fitted model, the_newdata) and
#'   produces predictions
#' @param error_fun Function. Has one arg: the test data, with fits added.
#' @param kfolds Integer. The number of folds.
kfold_cv <- function(data, estimator, predictor, error_fun, kfolds = 5) {
  n <- nrow(data)
  fold.labels <- sample(rep(1:kfolds, length.out = n))
  errors <- double(kfolds)
  for (fold in 1:kfolds) {
    test.rows <- fold.labels == fold
    train <- data[!test.rows, ]
    test <- data[test.rows, ]
    current_model <- estimator(train)
    test$.preds <- predictor(current_model, test)
    errors[fold] <- error_fun(test)
  }
  mean(errors)
}
```

::: {.solbox data-latex=""}
```{r q6-solution}
error_fun <- function(test) {
  sqrt(mean((test$`cases+14` - test$.preds)^2, na.rm = TRUE))
}

estimator1 <- function(train) {
  lm(`cases+14` ~ `cases+0` + `cases-7` + `cases-14`, data = train)
}

predictor1 <- function(model, newdata) {
  predict(model, newdata)
}

estimator2 <- function(train) {
  lm(log((`cases+14` / pop) * 100000) ~ 
       log((`cases+0`  / pop) * 100000) +
       log((`cases-7`  / pop) * 100000) +
       log((`cases-14` / pop) * 100000) +
       HA,
     data = train)
}

predictor2 <- function(model, newdata) {
  # Predict in log-per-capita space
  pred_log <- predict(model, newdata)
  
  # Back-transform to per-capita
  pred_pc <- exp(pred_log) 
  
  # Convert back to raw counts
  (pred_pc / 100000) * newdata$pop
}

set.seed(123)  # for reproducibility

cv1 <- kfold_cv(bccovid_train2, estimator1, predictor1, error_fun, kfolds = 10)
cv2 <- kfold_cv(bccovid_train2, estimator2, predictor2, error_fun, kfolds = 10)

cv1
cv2
```

a.  

b.  

c.  
:::

\clearpage

## (1 point)

Use your preferred model to predict the test set. a. (0.25 points) Plot
the actual values and the predictions, coloured by HA. Use lines for the
predictions and dots for the observations. b. (0.25 points) For each HA,
how accurate is the prediction on average (what is the mean absolute
error on the scale of cases)? c. (0.25 points) What about overall? d.
(0.25 points) How does this compare to the CV score you computed above?

::: {.solbox data-latex=""}
a.  

```{r q7-solution}
model_pref <- lm(`cases+14` ~ `cases+0` + `cases-7` + `cases-14`, data = bccovid_train)

pred_raw <- predict(model_pref, bccovid_test)

bccovid_test$.preds <- pred_raw


library(ggplot2)

ggplot(bccovid_test, aes(x = date, colour = HA)) +
  geom_point(aes(y = `cases+14`)) +             # actuals
  geom_line(aes(y = .preds)) +                  # predictions
  labs(title = "Predicted vs Actual Cases (Test Set)",
       y = "Weekly cases (2-week ahead)")


# MAE per HA
mae_by_HA <- bccovid_test %>%
  group_by(HA) %>%
  summarise(mae = mean(abs(`cases+14` - .preds), na.rm = TRUE))

# overall MAE
mae_overall <- sqrt(mean(abs(bccovid_test$`cases+14` - bccovid_test$.preds)^2, na.rm = TRUE))

mae_by_HA
mae_overall

```

b.  

c.  

d.  
:::

\clearpage

## (2 points)

a.  (0.25 points) Write the statistical model used for Question 1.3 (the
    log LM), and write the distributional assumption on the model
    residuals.

b.  (0.5 points) Assume we now modify the distribution on the residuals
    to be i.i.d. draws from [Laplace
    distribution](https://en.wikipedia.org/wiki/Laplace_distribution)
    with $\mu$ (location parameter) = 0 and $b$ (scale parameter) = 1. I
    claim that this model is more robust to outliers in the data. Why?
    Limit your answer to two sentences. (Hint: Plot the Laplace
    distribution and the Normal distribution side-by-side, and compare
    the tail behavior and probability density for large deviations from
    the mean.)

c.  (0.5 points) Derive the MLE for this model. This estimator does not
    have a closed form, so your final answer should be expressed as an
    optimization problem (i.e. argmax or argmin) over a summation, just
    like what we derived in class for logistic regression. Show your
    work for taking the derivative and setting it to zero, even though
    you cannot solve it explicitly.

d.  (0.25 points) What is the continuous risk function that, if we were
    to take an ERM perspective, yields the same estimator as the MLE?

e.  (0.5 points) Without actually fitting the model, give one (\~2
    sentence reason) why the Laplace model might perform better than the
    existing log LM for the COVID data. Give one (\~2 sentence reason)
    why the existing model might perform better. Base your answers on
    the graphs and metrics you have computed so far.

::: {.solbox data-latex=""}
a.  

b.  

c.  

d.  

e.  
:::
